---
title: "Host-layered GBM species distribution models for hemiparasite restoration suitability"
author: "Jordan"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

This workflow adapts Victor Felix's CEID "Machine learning SDM in R: Zero to
Pro(blems)" workshop into a **host-layered, two-stage** pipeline:

- **Stage 1 — Host SDMs:** model each host on bioclimatic + land-cover predictors.
- **Stage 2 — Hemiparasite SDMs:** model each parasite on the same environmental
  predictors **plus** the host suitability layer(s) from Stage 1.

It is written to run for **one focal parasite by default**, but every step is
list-driven, so adding more parasites (or more hosts per parasite) is just a
matter of adding entries in the configuration block below.

---

# 0. Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(readr)     # read GBIF download files
library(pROC)      # ROC / AUC
library(gbm)       # gradient boosting machine
library(dismo)
library(reshape2)
library(dplyr)
library(geodata)   # WorldClim + land cover
library(terra)     # raster / vector handling
library(tigris)    # US boundaries
library(sf)
library(ggplot2)
library(caret)     # train() wrapper for GBM

# Output folders (created if missing)
for (d in c("data", "data/species_data", "figures", "forecasts")) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}
```

---

# 1. Project configuration  *(SKELETON — fill in once species are chosen)*

> **You have not yet identified hosts or parasites, so leave these as placeholders
> for now.** When you are ready, uncomment one parasite (and its host[s]) to run
> the single-species case, then add more list entries to scale up. The loops in
> Stages 1–2 iterate over whatever you define here — no other code needs to change.

```{r config-species}
# ---- FOCAL PARASITE(S): name -> path to its GBIF occurrence download ----
# Start with ONE entry. Add more lines to model additional parasites.
parasite_species <- list(
  # "<parasite_species_1>" = "data/species_data/<parasite_1_gbif>.csv"
  # , "<parasite_species_2>" = "data/species_data/<parasite_2_gbif>.csv"
)

# ---- HOST(S) PER PARASITE: parasite -> named vector of host -> GBIF path ----
# Each parasite maps to one or more hosts. Hosts shared across parasites are
# only modeled once (Stage 1 caches them).
host_species <- list(
  # "<parasite_species_1>" = c(
  #   "<host_species_a>" = "data/species_data/<host_a_gbif>.csv"
  #   # , "<host_species_b>" = "data/species_data/<host_b_gbif>.csv"
  # )
)
```

```{r config-model}
# ---- GBM hyperparameters (from the workshop) ----
# NOTE: the workshop slides list n.trees = 360; the workshop code used 380.
# Adjust as needed for your data.
gbm_tune_grid <- expand.grid(
  interaction.depth = 2:3,   # tree depth
  n.trees           = 380,   # number of trees
  shrinkage         = 0.01,  # learning rate
  n.minobsinnode    = 50     # minimum observations per leaf
)

# Land-cover classes to include as predictors (ESA WorldCover fractions).
# Options: "trees","grassland","shrubs","cropland","built","bare",
#          "snow","water","wetland","mangroves","moss".
# Defaults below are sensible for Midwest grassland hemiparasites.
landcover_vars <- c("grassland", "cropland", "trees", "shrubs", "wetland")

# Reproducibility + train/test split
random_seed    <- 30
train_fraction <- 0.70
```

---

# 2. Study region

The workshop modeled the contiguous U.S. Your project targets the Midwest; the
contiguous-U.S. boundary is built first, with an **optional** Midwest filter you
can switch on once your extent is finalized.

```{r study-region}
# US states with geometries
us_states      <- tigris::states(year = 2023, cb = TRUE)
us_states_vect <- terra::vect(us_states)

# Drop non-contiguous states/territories
drop_codes    <- c("AK", "HI", "PR", "GU", "VI", "MP", "AS")
contiguous_us <- us_states_vect[!us_states_vect$STUSPS %in% drop_codes, ]

# ---- OPTION: restrict to Midwest states (uncomment to use) ----
# midwest_codes <- c("IL","IN","IA","KS","MI","MN","MO","NE","ND","OH","SD","WI")
# contiguous_us <- contiguous_us[contiguous_us$STUSPS %in% midwest_codes, ]

# Dissolve to a single study-region geometry
study_region_vect <- terra::aggregate(contiguous_us)

plot(study_region_vect, main = "Study region")
```

---

# 3. Environmental predictors (bioclimatic + land cover)

```{r worldclim}
# Bioclimatic rasters (~81 km² / 5 arc-min)
worldclim_rasters <- geodata::worldclim_global(var = "bio", res = 5, path = "data/")

# Simplified bioclim names (WorldClim BIO1–BIO19)
new_names <- c(
  "AnnMeanTemp", "MeanDiurnalRange", "Isotherm", "TempSeasonality",
  "MaxTempWarmM", "MinTempColdM", "TempAnnRange", "MeanTempWetQ",
  "MeanTempDryQ", "MeanTempWarmQ", "MeanTempColdQ", "AnnPrecip",
  "PrecipWetM", "PrecipDryM", "PrecipSeasonality", "PrecipWetQ",
  "PrecipDryQ", "PrecipWarmQ", "PrecipColdQ"
)
names(worldclim_rasters) <- new_names

# Crop + mask to the study region
worldclim_region <- terra::crop(worldclim_rasters, study_region_vect) |>
  terra::mask(study_region_vect)
```

```{r landcover}
# Land-cover fractions (geodata::landcover, per Victor's email).
# Each layer is the fraction of that class per cell; downloaded at 30 arc-sec
# and resampled onto the WorldClim grid so the stacks align.
landcover_list  <- lapply(landcover_vars, function(v) geodata::landcover(var = v, path = "data/"))
landcover_stack <- terra::rast(landcover_list)
names(landcover_stack) <- paste0("lc_", landcover_vars)

landcover_region <- terra::crop(landcover_stack, study_region_vect) |>
  terra::resample(worldclim_region) |>   # match resolution/grid of climate layers
  terra::mask(study_region_vect)

# Final predictor stack shared by every model (Stage 2 appends host layers to this)
env_predictors <- c(worldclim_region, landcover_region)
names(env_predictors)
```

---

# 4. Reusable pipeline function

`run_gbm_sdm()` encapsulates the entire per-species workshop workflow:
load GBIF occurrences → reduce to binary presence per grid cell → sample
background → 70/30 split → train a 3-fold cross-validated GBM → evaluate with
AUC → forecast a suitability raster. It accepts **any** predictor stack, so the
same function serves both the host stage and the host-layered parasite stage.

```{r pipeline-function}
run_gbm_sdm <- function(species_name,
                        occurrence_path,
                        predictors,
                        region_vect,
                        hyperparams    = gbm_tune_grid,
                        bg_multiplier  = 1,      # background points = bg_multiplier * presence
                        train_frac     = train_fraction,
                        seed           = random_seed,
                        save_figures   = TRUE,
                        out_dir        = "forecasts") {

  message("=== GBM SDM: ", species_name, " ===")
  safe_name <- gsub("[^A-Za-z0-9]+", "_", species_name)

  ## --- 1. Load GBIF occurrences (standard tab-delimited GBIF export) ---
  occ_raw <- readr::read_delim(occurrence_path, delim = "\t",
                               escape_double = FALSE, col_names = TRUE,
                               na = c("", "NA"), show_col_types = FALSE)

  occ <- data.frame(longitude = occ_raw$decimalLongitude,
                    latitude  = occ_raw$decimalLatitude,
                    id        = 1) |>
    stats::na.omit()

  occ_vect <- terra::vect(occ, geom = c("longitude", "latitude"), crs = "EPSG:4326")
  occ_vect <- terra::crop(occ_vect, region_vect)

  ## --- 2. Rasterize to the predictor grid; reduce to BINARY presence per pixel ---
  ref_layer  <- predictors[[1]]
  occ_raster <- terra::rasterize(occ_vect, ref_layer, field = "id",
                                 fun = "count", background = 0)
  occ_raster[occ_raster > 0] <- 1

  presence_df <- terra::as.data.frame(occ_raster, xy = TRUE)
  colnames(presence_df) <- c("longitude", "latitude", "Label")
  presence_df <- dplyr::filter(presence_df, Label == 1)
  presence_pts <- terra::vect(presence_df, geom = c("longitude", "latitude"),
                              crs = "EPSG:4326")

  ## --- 3. Extract predictors at presence points ---
  presence_env <- terra::extract(predictors, presence_pts)
  presence_env <- subset(presence_env, select = -c(ID))
  presence_env$Label <- 1
  presence_env <- stats::na.omit(presence_env)

  ## --- 4. Random background points (same count as presences by default) ---
  set.seed(seed)
  n_bg <- round(bg_multiplier * nrow(presence_env))
  bg_xy <- terra::spatSample(predictors, size = n_bg, method = "random",
                             na.rm = TRUE, xy = TRUE) |>
    dplyr::mutate(Label = 0)
  background_env <- subset(bg_xy, select = -c(x, y))
  background_env <- stats::na.omit(background_env)

  ## --- 5. 70/30 train-test split (presence and background split separately) ---
  set.seed(seed)
  p_idx <- sample(seq_len(nrow(presence_env)),   size = floor(train_frac * nrow(presence_env)))
  b_idx <- sample(seq_len(nrow(background_env)), size = floor(train_frac * nrow(background_env)))

  train_data <- rbind(presence_env[p_idx, ],  background_env[b_idx, ])
  test_data  <- rbind(presence_env[-p_idx, ], background_env[-b_idx, ])

  # Target as a factor with caret-friendly level names
  to_factor <- function(df) {
    df$Label <- factor(df$Label, levels = c(0, 1), labels = c("Background", "Presence"))
    df
  }
  train_data <- to_factor(train_data)
  test_data  <- to_factor(test_data)

  feature_cols <- setdiff(names(train_data), "Label")  # works for any # of predictors

  ## --- 6. Train GBM (3-fold CV, ROC-selected) ---
  set.seed(11)
  gbm_model <- caret::train(
    x = train_data[, feature_cols],
    y = train_data$Label,
    method   = "gbm",
    metric   = "ROC",
    trControl = caret::trainControl(
      method          = "cv",
      number          = 3,
      classProbs      = TRUE,
      summaryFunction = caret::twoClassSummary
    ),
    tuneGrid = hyperparams,
    verbose  = FALSE
  )
  train_auc <- mean(gbm_model$resample$ROC)

  ## --- 7. Test-set AUC ---
  test_probs <- predict(gbm_model, test_data[, feature_cols], type = "prob")[, "Presence"]
  test_roc   <- pROC::roc(test_data$Label, test_probs, quiet = TRUE)
  test_auc   <- as.numeric(pROC::auc(test_roc))
  message(sprintf("  train AUC = %.3f | test AUC = %.3f", train_auc, test_auc))

  if (save_figures) {
    roc_df <- data.frame(fpr = 1 - test_roc$specificities, tpr = test_roc$sensitivities)
    test_plot <- ggplot(roc_df, aes(fpr, tpr)) +
      geom_line(color = "blue", linewidth = 1.1) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
      labs(title = paste("GBM Testing ROC:", species_name),
           subtitle = paste("AUC =", round(test_auc, 3)),
           x = "1 - Specificity", y = "Sensitivity") +
      theme_minimal(base_size = 13)
    ggsave(file.path("figures", paste0(safe_name, "_test_ROC.jpg")),
           test_plot, width = 4, height = 4, dpi = 300)

    # Feature importance
    imp <- caret::varImp(gbm_model)$importance
    imp$Percentage <- 100 * imp$Overall / sum(imp$Overall)
    imp$Variable   <- rownames(imp)
    imp_plot <- ggplot(imp, aes(reorder(Variable, Percentage), Percentage)) +
      geom_segment(aes(xend = Variable, y = 0, yend = Percentage), color = "blue") +
      geom_point(color = "blue", size = 2.5) + coord_flip() +
      labs(title = paste("GBM feature importance:", species_name),
           x = "Model features", y = "Percentage (%)") +
      theme_minimal(base_size = 12)
    ggsave(file.path("figures", paste0(safe_name, "_importance.jpg")),
           imp_plot, width = 5, height = 5, dpi = 300)
  }

  ## --- 8. Forecast a suitability raster over the study region ---
  suitability <- terra::predict(predictors, gbm_model, type = "prob", na.rm = TRUE)
  suitability <- suitability[["Presence"]]
  names(suitability) <- safe_name
  suitability <- terra::mask(suitability, region_vect)
  terra::writeRaster(suitability,
                     filename = file.path(out_dir, paste0(safe_name, "_SDM.tif")),
                     overwrite = TRUE)

  ## --- Return everything the next stage needs ---
  list(species     = species_name,
       model       = gbm_model,
       train_auc   = train_auc,
       test_auc    = test_auc,
       suitability = suitability)
}
```

---

# 5. Stage 1 — Host SDMs  *(skeleton; set `eval=TRUE` once hosts are defined)*

Models each host once and stores its suitability layer for use in Stage 2.

```{r stage1-hosts, eval=FALSE}
host_models      <- list()  # fitted host GBMs
host_suitability <- list()  # host -> suitability SpatRaster

for (parasite in names(host_species)) {
  for (host in names(host_species[[parasite]])) {
    if (!is.null(host_suitability[[host]])) next  # already modeled (shared host)

    res <- run_gbm_sdm(
      species_name    = host,
      occurrence_path = host_species[[parasite]][[host]],
      predictors      = env_predictors,        # climate + land cover
      region_vect     = study_region_vect,
      hyperparams     = gbm_tune_grid
    )

    host_models[[host]]      <- res$model
    host_suitability[[host]] <- res$suitability
  }
}
```

---

# 6. Stage 2 — Hemiparasite SDMs (host-layered)  *(skeleton; set `eval=TRUE` to run)*

For each parasite, the predictor stack = environmental predictors **+** a single
**averaged host suitability** layer (the cell-wise mean of its host suitability
rasters from Stage 1). This matches the one-pager methodology and is the
host-layered step that distinguishes the model from a climate-only SDM. With one
host the mean is just that host's layer; with several it collapses them into one
combined "host availability" predictor.

```{r stage2-parasites, eval=FALSE}
parasite_models      <- list()
parasite_suitability <- list()

for (parasite in names(parasite_species)) {

  # Gather this parasite's host suitability layers (one raster per host)
  hosts       <- names(host_species[[parasite]])
  host_layers <- terra::rast(host_suitability[hosts])

  # Average across hosts into a SINGLE combined host-suitability predictor,
  # matching the one-pager ("averaged host suitability layers").
  # (Cell-wise mean across host layers; with one host this is just that layer.)
  host_suit_mean <- terra::app(host_layers, fun = "mean", na.rm = TRUE)
  names(host_suit_mean) <- "hostsuit_mean"

  # Host-layered predictor stack: climate + land cover + averaged host suitability
  predictors_stage2 <- c(env_predictors, host_suit_mean)

  res <- run_gbm_sdm(
    species_name    = parasite,
    occurrence_path = parasite_species[[parasite]],
    predictors      = predictors_stage2,       # climate + land cover + host suitability
    region_vect     = study_region_vect,
    hyperparams     = gbm_tune_grid
  )

  parasite_models[[parasite]]      <- res$model
  parasite_suitability[[parasite]] <- res$suitability

  # Quick map
  plot(res$suitability,
       main = paste0(parasite, " suitability (test AUC = ", round(res$test_auc, 3), ")"),
       col  = colorRampPalette(c("white", "darkgreen"))(100), range = c(0, 1))
  plot(study_region_vect, add = TRUE)
}
```

---

# 7. Restoration suitability overlay  *(placeholder for Aim 3)*

Once parasite suitability rasters exist, overlay them with restoration-site
coordinates to rank candidate sites. Fill in your sites table when ready.

```{r restoration-overlay, eval=FALSE}
# restoration_sites <- readr::read_csv("data/restoration_sites.csv")  # site, longitude, latitude, ...
# sites_vect <- terra::vect(restoration_sites, geom = c("longitude", "latitude"), crs = "EPSG:4326")
#
# for (parasite in names(parasite_suitability)) {
#   restoration_sites[[paste0("suit_", parasite)]] <-
#     terra::extract(parasite_suitability[[parasite]], sites_vect)[, 2]
# }
#
# # Rank sites by predicted suitability (and, if desired, host suitability)
# ranked_sites <- dplyr::arrange(restoration_sites, dplyr::desc(.data[[paste0("suit_", names(parasite_suitability)[1])]]))
# head(ranked_sites)
```
