---
title: "Machine learning species distribution models in R: Zero to pro(blems)"
author: "Victor Felix (victorfelix@uga.edu)"
date: "2026-04-07"
output:
  pdf_document: default
  html_document: default
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

Let's start loading some libraries

```{r}
#install.packages("pROC") 
library(pROC)
#install.packages("gbm") 
library(gbm)
#install.packages("dismo") 
library(dismo)
#install.packages("reshape2") 
library(reshape2)
#install.packages("dplyr") 
library(dplyr)
#install.packages("geodata") 
library(geodata)
#install.packages("terra") 
library(terra)
#install.packages("tigris") 
library(tigris)
#install.packages("sf") 
library(sf)
#install.packages("ggplot2") 
library(ggplot2)
#install.packages("caret") 
library(caret)
```

Here I get the shapefile of the contiguous U.S.

```{r}
# US states dataframe with geometries 
us_states <- tigris::states(year = 2023, cb = TRUE)

# Convert to terra SpatVector
us_states_vect <- terra::vect(us_states)

# Select only the contiguous US
contiguous_us <- us_states_vect[!us_states_vect$STUSPS %in% c("AK", "HI", "PR", "GU", "VI", "MP", "AS"), ]

# AK:	Alaska, HI:	Hawaii, PR:	Puerto Rico, GU:	Guam, VI:	U.S. Virgin Islands, MP:	Northern Mariana Islands, AS:	American Samoa

# Dissolve all states into one geometry
contiguous_us_vect <- terra::aggregate(contiguous_us)

# Save figure for the presentation
png("figures/contiguous_us_map.png", width = 1000, height = 800, res = 250)
plot(contiguous_us_vect, main="Contiguous United States")
dev.off()

# Plot contiguous US
plot(contiguous_us_vect)

```

# Downloading WorldClim rasters

```{r}
# Downloading WorldClim rasters
worldclim_rasters<-geodata::worldclim_global(var = "bio", # variables to download
                                 res = 5, # 5 arc-minute or ~81 km²
                                 path = "data/") # path were data is saved
# Plot rasters
plot(worldclim_rasters)
```
Renaming WorldClim data based on simplified variables names.

https://www.worldclim.org/data/bioclim.html#google_vignette

```{r}
# New simplified names
new_names <- c(
  "AnnMeanTemp", "MeanDiurnalRange", "Isotherm", "TempSeasonality",
  "MaxTempWarmM", "MinTempColdM", "TempAnnRange", "MeanTempWetQ",
  "MeanTempDryQ", "MeanTempWarmQ", "MeanTempColdQ", "AnnPrecip",
  "PrecipWetM", "PrecipDryM", "PrecipSeasonality", "PrecipWetQ",
  "PrecipDryQ", "PrecipWarmQ", "PrecipColdQ"
)

# Rename rasters
base::names(worldclim_rasters) <- new_names

plot(worldclim_rasters)

```
Cropping and masking the stacked raster to the size of the contiguous U.S.

```{r}
# Defining some coordinates for contiguous US
us_extent <- ext(-125, -66.5, 24.5, 49.5)  # approximate lat/lon 

# Cropping and masking the rasters
worldclim_rasters_us <- terra::crop(worldclim_rasters, contiguous_us_vect)%>%
  mask(contiguous_us_vect)

# Save figure for the presentation
png("figures/worldclim_rasters.png", width = 1000, height = 600, res = 150)
plot(worldclim_rasters_us)
dev.off()

# Plot
plot(worldclim_rasters_us)
```

Download American Green Treefrog data.

https://www.gbif.org/species/10810025

Dryophytes cinereus (Schneider, 1799)

GBIF.org (11 December 2025) GBIF Occurrence Download https://doi.org/10.15468/dl.kg3eax

```{r }
# Loading the dataset - this is standard for GBIF data on this format

AGTF_raw <- readr::read_delim("data/species_data/american_green_tree_frog.csv", delim = "\t", escape_double = FALSE, 
    col_names = TRUE, na = c("", "NA"))

head(AGTF_raw)
```

Getting the coordinates from the GBIF occurrences.

```{r}
# Create a data frame using latitudes and longitudes
# id == 1 means presence
AGTF_occurences <- base::data.frame("latitude" = AGTF_raw$decimalLatitude, 
                            "longitude" = AGTF_raw$decimalLongitude,
                            "id"=1)

# Remove NAs - places without Latitude or Longitude data
AGTF_occurences <- stats::na.omit(AGTF_occurences)

head(AGTF_occurences)
```

Plotting these occurrences in a map so we can see.
We will use the contiguous US vector to select only occurrences in the study region. 

```{r}
# Create a vector with these locations (shapefile points)
AGTF_occurences <- terra::vect(AGTF_occurences, geom = c("longitude", "latitude"))

# Plotting data and contiguous US
plot(AGTF_occurences, main="Green Treefrog occurences from GBIF")
plot(contiguous_us_vect, add=TRUE)
```

Keep only occurrences in the contiguous U.S.

!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
# this step take some time to run  
# time for 1 or 2 questions
!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

```{r}
# Crop the raster using the contiguous U.S. area
AGTF_occurences_US <- terra::crop(AGTF_occurences, contiguous_us_vect)

# Saving figure for presentation
png("figures/occurrences_points.png", width = 1000, height = 600, res = 150)
plot(AGTF_occurences_US, main="Green Treefrog occurrences")
plot(contiguous_us_vect ,add=TRUE)
dev.off()

plot(AGTF_occurences_US, main="Green Treefrog occurrences")
plot(contiguous_us_vect ,add=TRUE)
```

Transforming the occurrences (points in a vector format) to a raster format.
For this, we use a WorldClim raster as reference and create a new raster based on it. 
The new raster will have on each pixel cell the sum of Green Treefrog occurrences in that area.

```{r}
# Creating the new raster based on a WorldClim raster and
# Counting the number of occurrences in each raster cell
# Empty pixels are set to 0.
AGTF_occurences_raster <- terra::rasterize(AGTF_occurences_US, field="id", worldclim_rasters_us$AnnMeanTemp, fun = "count", background = 0, filename="data/AGTF_count.tif", overwrite=TRUE)

# Save figure for presentation
png("figures/occurrences_raster.png", width = 1000, height = 650, res = 150)

plot(AGTF_occurences_raster,
     main = "Green Treefrog: Number of occurrences per raster pixel ~81 km²",
     sub = "Number of occurrences per raster cell ~81 km²")

plot(contiguous_us_vect, border = "white", add = TRUE)

dev.off()

# Plot figure
plot(AGTF_occurences_raster,
     main = "Green Treefrog: Number of occurrences per raster pixel ~81 km²",
     sub = "Number of occurrences per raster pixel ~81 km²")

plot(contiguous_us_vect, border = "white", add = TRUE)

```

It is better to transform this raster to a dataframe and plot with ggplot
This will help us to visualize the number of Green Treefrogs occurrences

```{r}
# Convert raster to data frame to plot with ggplot
AGTF_occurences_df <- base::as.data.frame(AGTF_occurences_raster, xy = TRUE)
colnames(AGTF_occurences_df) <- c("x", "y", "value")

# Filter dataframe to keep only values > 0
AGTF_occurences_df <- AGTF_occurences_df[AGTF_occurences_df$value > 0, ]


AGTF_occurences_plot<-ggplot() +
  geom_sf(data = sf::st_as_sf(contiguous_us_vect), # convert to sf to plot
          fill = "white", color = "black", linewidth = 0.2) +
  geom_point(
    data = AGTF_occurences_df,
    aes(x = x, y = y, size = value, color = value),
    alpha = 0.6
  ) +
  scale_size_continuous(
  range = c(1, 10),
  name = "",
  breaks = c(1, 50, 100, 200, 300)
) +
  scale_color_gradient(low = "lightgreen", high = "darkgreen", name = "") +
  labs(
    title = "Green Treefrog occurrences",
    subtitle = "Number of occurences per ~81 km² (raster pixel size)",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 19),
    plot.subtitle = element_text(size = 14),
    legend.position = "bottom"
  )

AGTF_occurences_plot

ggsave("figures/density_plot.jpg", AGTF_occurences_plot, width = 8,
  height = 6,  dpi = 300)

# visualize the dataframe
head(AGTF_occurences_df)
```

Now, for modelling purposes, we will assume that locations with at least one occurrence per pixel represent a presence point of the Green Treefrog species.


```{r}

# Set pixels values with at least one occurrence to 1 (presence)
AGTF_occurences_final_raster <- AGTF_occurences_raster
AGTF_occurences_final_raster[AGTF_occurences_final_raster > 0] <- 1

# Convert raster to dataframe
AGTF_occurences_final_df <- base::as.data.frame(AGTF_occurences_final_raster, xy = TRUE)

# rename columns
colnames(AGTF_occurences_final_df) <- c("longitude", "latitude", "Label")

# filter only Label == 1 (presence points)
AGTF_presence_df <- AGTF_occurences_final_df %>%
  filter(Label == 1)

# Transform to vector of points
AGTF_presence_points <- terra::vect(
  AGTF_presence_df,
  geom = c("longitude", "latitude"),
  crs = "EPSG:4326"
)


```

Let's plot for easier visualization. 

```{r}
# Save the figure for presentation
png("figures/presence_points.png", width = 1000, height = 600, res = 150)
plot(contiguous_us_vect,
     xlab = "Longitude", ylab = "Latitude",
     main = "Presence points")
points(AGTF_presence_df$longitude, AGTF_presence_df$latitude,
       pch = 16, cex = 0.2, col = "darkgreen")
dev.off()

# plotting presence points
plot(contiguous_us_vect,
     xlab = "Longitude", ylab = "Latitude",
     main = "Presence points")
points(AGTF_presence_points,
       pch = 16, cex = 0.2, col = "darkgreen")

```

######
Back to slides
#####

Now we will use the vector of points to extract the climatic data for each presence point.

```{r}
presence_points <- terra::extract(worldclim_rasters_us, AGTF_presence_points)

# Drop ID column
presence_points <- base::subset(presence_points, select = -c(ID))
# Create a Label column with 1 as presence
presence_points$Label <- 1

# visulize dataframe
head(presence_points)

# numbr of rows
NROW(presence_points)
```

Now we randomly select the same number of background points as presence points, and sample the climate data from each of these points.

```{r}
set.seed(30)
# Sampling random background points 
x_y_background_points <- spatSample(worldclim_rasters_us, size = NROW(presence_points), method = "random", na.rm = TRUE, xy=TRUE)%>%
  mutate(Label = 0)

# save the plot
png("figures/background_points.png", width = 1000, height = 600, res = 150)
plot(x_y_background_points$x, x_y_background_points$y,
     pch = 16, cex = 0.3,
     xlab = "Longitude", ylab = "Latitude",
     main = "Background points")
plot(contiguous_us_vect,add=TRUE)
dev.off()

#Plot 
plot(x_y_background_points$x, x_y_background_points$y,
     pch = 16, cex = 0.4,
     xlab = "Longitude", ylab = "Latitude",
     main = "Background points")
plot(contiguous_us_vect,add=TRUE)

# sample climate data from these points
background_points <- subset(x_y_background_points, select = -c(x, y))

head(background_points)
```

Get 70% of the data for training and 30% for testing.

```{r}
# Randomly sample 70% of the background indices
sample_indices <- base::sample(seq_len(nrow(background_points)), size = 0.7 * nrow(background_points))
# Create a sampled dataframe with 70% of the data
bg_70 <- background_points[sample_indices, ]

# Randomly sample 70% of the presence row indices
sample_indices <- base::sample(seq_len(nrow(presence_points)), size = 0.7 * nrow(presence_points))
# Create a sampled dataframe with 70% of the data
presence_70 <- presence_points[sample_indices, ]

#rbind 70% of presence and background for training data
train.data.70<-base::rbind(presence_70,bg_70)

# Get the 30% data left for presence and background data 
test_presence_30 <- presence_points[-sample_indices, ]
test_bg_30 <- background_points[-sample_indices, ]

#rbind the 30% left of presence and background for testing
test.data.30<-rbind(test_bg_30,test_presence_30)

```

Model setup and tuning

```{r}
############
set.seed(11)

# Choosing training data
train.data <- train.data.70
train.data$Label <- as.factor(train.data$Label) # Define target column as a factor
levels(train.data$Label) <- c("Background", "Presence") # rename 0s and 1s 

#################
# GBM model setup
#################

gbm_model <- train(
  x = train.data[, 1:19], # 19 first columns with climate data
  y = train.data[, 20], # presence or background points
  method = "gbm", # Gradient boosting machine
  metric = "ROC", # ROC for model selection
  trControl = trainControl(
    method = "cv", # cross-validation
    number = 3, # three folds
    classProbs = TRUE, # compute predicted probabilities 
    summaryFunction = twoClassSummary # computes ROC, sensitivity and specificity
  ),
  tuneGrid = expand.grid(
    interaction.depth = 2:3, # trees depth
    n.trees = 380, # Total number of trees
    shrinkage = 0.01, # learning rate
    n.minobsinnode = 50 # min observations in a leaf
  ),
  verbose = TRUE # show models
)

# Print the mean cross-validation AUC
cat("AUC:", mean(gbm_model$resample$ROC), "\n")

```

Plotting GBM training ROC

```{r}
# Get predicted probabilities 
pred_probs <- predict(gbm_model, newdata = train.data[, 1:20], type = "prob")[, "Presence"]

# Compute ROC curve
roc_obj <- roc(response = train.data$Label, predictor = pred_probs)

# Create a data frame for plotting
roc_df <- data.frame(
  tpr = roc_obj$sensitivities,
  fpr = 1 - roc_obj$specificities
)

# Plot 
train_plot <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "blue", size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  labs(
    title = paste("GBM Training ROC"),
    subtitle = paste("AUC =", round(mean(gbm_model$resample$ROC), 3)),
    x = "1 - Specificity",
    y = "Sensitivity"
  ) +
  theme_minimal(base_size = 14)

#####
ggsave("figures/train_ROC.jpg", plot = train_plot, width = 3, height = 3, dpi = 300)

print(train_plot)
```

GBM testing ROC

```{r}
# Loading test data and transforming labels to factor
test.data <- test.data.30
test.data$Label <- as.factor(test.data$Label) # set as a factor
levels(test.data$Label) <- c("Background", "Presence") # rename 0s and 1s

# Forecasts for test data using our trained model
predictions <- predict(gbm_model, test.data[, 1:20], type = "prob")

# Extract probabilities
prob_positive <- predictions$Presence

# Run the ROC curve calculation
roc_curve <- roc(test.data$Label, prob_positive)

# Calculate the AUC
auc_value <- auc(roc_curve)

# Create dataframe for plotting
roc_df <- data.frame(
  fpr = 1 - roc_curve$specificities,
  tpr = roc_curve$sensitivities
)

# Plot
test_plot <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "blue", size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  labs(
    title = paste("GBM Testing ROC"),
    subtitle = paste("AUC =", round(auc(roc_curve), 3)),
    x = "1 - Specificity",
    y = "Sensitivity"
  ) +
  theme_minimal(base_size = 14)

test_plot

# Save the ggplot
ggsave("figures/test_ROC.jpg", plot = test_plot, width = 3, height = 3, dpi = 300)
```

Plotting GBM variable importance.

```{r}
set.seed(11)
# Calculate variable importance
importance <- varImp(gbm_model)
# Extract the importance values as a data frame
feature_importance <- as.data.frame(importance$importance)
# Add a column for the percentage importance
feature_importance$Percentage <- (feature_importance$Overall / sum(feature_importance$Overall)) * 100
# Sort by percentage importance 
feature_importance <- feature_importance[order(-feature_importance$Percentage), ]
# Convert rownames to a column for easier plotting
feature_importance$Variable <- rownames(feature_importance)

# Plot
importance_plot <- ggplot(feature_importance, aes(x = reorder(Variable, Percentage), y = Percentage)) +
  geom_segment(aes(xend = Variable, y = 0, yend = Percentage), color = "blue") +
  geom_point(color = "blue", size = 3) +
  coord_flip() +
  labs(
    title = "GBM Feature importance",
    x = "Model features",
    y = "Percentage (%)"
  ) +
  theme_minimal()

importance_plot

# Save the ggplot
ggsave("figures/feature_importance.jpg", plot = importance_plot, width = 4, height = 4, dpi = 300)
```

Forecasting the contiguous U.S.

```{r}

map_pred <- terra::predict(object=worldclim_rasters_us, model=gbm_model,filename= "forecasts/US_SDM.tif",overwrite=TRUE,type = "prob")

# load the forecasts
model_pred<-rast("forecasts/US_SDM.tif")
# mask the forecasts
model_pred_masked<-mask(model_pred,contiguous_us_vect)


# save plot for presentation
png("figures/GreenTreefrogSDM.png", width = 1000, height = 600, res = 150)
plot(model_pred_masked$Presence, main="Green Treefrog Species Distribution Model", xlab="",
     col = colorRampPalette(c("white", "darkgreen"))(100),
     range = c(0, 1))
plot(contiguous_us_vect, add=TRUE)
dev.off()

# plot figure
plot(model_pred_masked$Presence, main="Green Treefrog Species Distribution Model", xlab="",
     col = colorRampPalette(c("white", "darkgreen"))(100),
     range = c(0, 1))
plot(contiguous_us_vect, add=TRUE)

```
